---
title: "exploratory_data_analysis"
author: "Rutika Avinash Kadam"
date: "2025-04-08"
output: html_document
---
```{r}
library(plotly)
library(ggplot2)
library(dplyr)
library(corrplot)
```
```{r}
clean_data <- read.csv("D:\\Projects\\Statistical Computing Project\\GlobalHealthStat\\data\\cleaned_data_for_EDA.csv")
head(clean_data)
```
```{r}
str(clean_data)
```
```{r}

#Find the number of unique values in each column
unique_values <- sapply(clean_data, function(x) length(unique(x)))

#Display the result
print(unique_values)
```
```{r}
#Separating the numeric and categoric features
numeric_cols <- clean_data %>% select_if(is.numeric)
categoric_cols <- clean_data %>% select_if(~!is.numeric(.))

cat("We have ",ncol(categoric_cols)," categorical columns.","\n","\n")
colnames(categoric_cols)
cat("\n")
cat("We have ",ncol(numeric_cols)," numerical columns.","\n","\n")
colnames(numeric_cols)
```
```{r}
#Identify categorical columns (factor or character columns)
categoric_cols <- sapply(clean_data, function(x) is.factor(x) | is.character(x))
categoric_cols <- names(categoric_cols[categoric_cols])

#Print unique values for each categorical column
for (col in categoric_cols) {
  unique_vals <- unique(clean_data[[col]])
  num_unique <- length(unique_vals)
  cat("Unique values in column", col,"are ",num_unique, ":\n")
  print(unique(clean_data[[col]]))
  cat("\n")
}
```
```{r}
#categoric_cols <- c("country", "disease_name", "disease_category", "age_group", 
#                    "gender", "treatment_type", "availability_of_vaccines_treatment")
# Loop through each column to display the value-to-number mapping
#for (col in categoric_cols) {
#  cat("Label Encoding for", col, ":\n")
  
#  if (col == "age_group") {
    # Define custom order for age_group
#    custom_levels <- c("child", "young adult", "adult", "senior")
#   encoding <- factor(clean_data[[col]], levels = custom_levels)
#  } else {
#    encoding <- factor(clean_data[[col]])
#  }
  
  # Create the value-to-number mapping
#  value_to_number <- data.frame(Value = levels(encoding), Encoded = 0:(length(levels(encoding)) - 1))
  
#  print(value_to_number)
#  cat("\n")
#}
```
```{r}
#Implementing Label Encoding for all categorical veriables
# Apply Label Encoding to each categorical column, starting from 0
# Define custom order for 'age_group' column
#age_group_order <- c("Child", "Young Adult", "Adult", "Senior")

# Loop through each column to apply custom encoding and print the mapping for 'age_group'
#for (col in categoric_cols) {
#  if (col == "age_group") {
    # Apply custom encoding for 'age_group' column
#    clean_data[[col]] <- factor(clean_data[[col]], levels = age_group_order)
#    clean_data[[col]] <- as.numeric(clean_data[[col]]) - 1  # Start encoding from 0
#  } else {
    # Apply label encoding for other columns
#    clean_data[[col]] <- as.numeric(factor(clean_data[[col]])) - 1
#  }
#}
#head(clean_data)
```
```{r}
# Define your numeric column names
# Define your numeric column names
numeric_cols <- c("year", "prevalence_rate", "incidence_rate", "mortality_rate", 
                  "population_affected", "healthcare_access", "doctors_per_1000", 
                  "hospital_beds_per_1000", "average_treatment_cost_usd", "recovery_rate", 
                  "dal_ys", "improvement_in_5_years", "per_capita_income_usd", 
                  "education_index", "urbanization_rate")

# Initialize vectors
columns_with_outliers <- c()
outlier_counts <- c()

# Loop over each numeric column to detect outliers and count them
for (col in numeric_cols) {
  # Extract the column and remove NA values
  x <- clean_data[[col]]
  x <- x[!is.na(x)]
  
  # Check if the column has enough data to compute quartiles
  if (length(x) < 4) next
  
  # Calculate Q1, Q3, and the IQR
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR_val <- IQR(x)
  
  # Calculate lower and upper bounds
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  
  # Find outlier indices
  outlier_indices <- which(x < lower_bound | x > upper_bound)
  
  # If any outlier exists, record it
  if (length(outlier_indices) > 0) {
    columns_with_outliers <- c(columns_with_outliers, col)
    outlier_counts[col] <- length(outlier_indices)
  }
}

# Create a data frame summarizing the outlier counts
outlier_summary <- data.frame(
  Column = names(outlier_counts),
  Outlier_Count = as.integer(outlier_counts)
)

# Print the summary sorted by count
outlier_summary <- outlier_summary[order(-outlier_summary$Outlier_Count), ]
print(outlier_summary)


```
```{r}
# Assuming your dataset is named 'df'
# Define the variable names and corresponding titles for the plots
variables <- c("average_treatment_cost_usd", "education_index", "per_capita_income_usd")
titles <- c("Average Treatment Cost (USD)",
            "Education Index",
            "Per Capita Income (USD)")

# Create an empty list to store the interactive plots
plot_list <- list()

# Loop over each variable to create individual box plots
for (i in seq_along(variables)) {
  p <- ggplot(clean_data, aes_string(x = "''", y = variables[i])) +
    geom_boxplot(fill = "pink", color = "black", 
                 outlier.colour = "red", outlier.shape = 16) +
    labs(title = titles[i],
         x = "",
         y = titles[i]) +
    theme_minimal()
  
  # Convert the ggplot object into an interactive plotly object and store it
  plot_list[[i]] <- ggplotly(p)
}

# Arrange the plots in one row (nrows = 1) while keeping each plot's y-axis independent
final_plot <- subplot(plot_list, nrows = 1, shareY = FALSE, titleX = TRUE, titleY = TRUE) %>% 
  layout(
    title = list(
      text = "Box Plots for Avg Treatment Cost, Education Index, & Per Capita Income(USD)",
      x = 0.5  # Center the main title
    )
  )

# Display the final interactive plot
final_plot
```
```{r}
#Removing Outliers-replacing outliers with median of that column
#Define target columns where you want to replace outliers with median
target_cols <- c("average_treatment_cost_usd", "mortality_rate", "prevalence_rate")

#Loop through each column
for (col in target_cols) {
  #Extract the column vector
  x <- clean_data[[col]]
  
  #Skip if all values are NA or fewer than 4 non-NA values
  if (sum(!is.na(x)) < 4) next
  
  #Compute Q1, Q3, and IQR
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  
  #Define bounds
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  
  #Calculate the median (excluding NA)
  med <- median(x, na.rm = TRUE)
  
  #Replace outliers with median
  x[x < lower_bound | x > upper_bound] <- med
  
  #Assign the cleaned column back
  clean_data[[col]] <- x
}
```
```{r}
#Check for null/missing values
clean_data[clean_data == ""] <- NA  

# Function to calculate missing values
missing_summary <- data.frame(
  Column_Name = names(clean_data),
  Missing_Count = sapply(clean_data, function(x) sum(is.na(x))),
  Missing_Percentage = sapply(clean_data, function(x) round(sum(is.na(x)) / length(x) * 100, 2))
)
# Print missing value summary
print(missing_summary)
```
```{r}
#Saving final without outliers and label encoding
write.csv(clean_data, file = "D:\\Projects\\Statistical Computing Project\\GlobalHealthStat\\data\\final_data1.csv", row.names = FALSE)
```
```{r}
final_data <- read.csv("D:\\Projects\\Statistical Computing Project\\GlobalHealthStat\\data\\final_data1.csv")
head(final_data)
```

```{r}
library(gridExtra)
library(ggpubr)  


# Sort numeric columns alphabetically (optional)
numeric_cols <- sort(c("prevalence_rate", "incidence_rate", "mortality_rate", 
                       "population_affected", "healthcare_access", "doctors_per_1000", 
                       "hospital_beds_per_1000", "average_treatment_cost_usd", 
                       "recovery_rate", "dal_ys", "improvement_in_5_years", 
                       "per_capita_income_usd", "education_index", "urbanization_rate"))

# Create list to store plots
plot_list <- list()

# Create individual plots with styling
for (col in numeric_cols) {
  p <- ggplot(final_data, aes_string(x = col)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "pink", alpha = 0.5) +
    geom_density(color = "red", size = 0.8) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.title.x = element_text(size = 10, face = "plain"), 
      axis.title.y = element_text(size = 10, face = "plain"),
      axis.text.x = element_text(size = 9, face = "plain"),   # X ticks
      axis.text.y = element_text(size = 9, face = "plain") 
    ) +
    labs(
      title = paste("Histogram & Density of", gsub("_", " ", col)),
      x = gsub("_", " ", col),
      y = "Density"
    )
  
  plot_list[[length(plot_list) + 1]] <- p
}

# Display 2 plots vertically per window
for (i in seq(1, length(plot_list), by = 2)) {
  plots_to_show <- plot_list[i:min(i+1, length(plot_list))]
  do.call(grid.arrange, c(plots_to_show, ncol = 1))
}
```
```{r}
library(ggplot2)
library(dplyr)



# Basic improved scatterplot
ggplot(final_data, aes(x = prevalence_rate, y = mortality_rate, color = disease_name)) +
  geom_point(alpha = 0.5, size = 2) +
  theme_minimal() +
  scale_color_viridis_d(option = "plasma") +  # nicer color palette
  labs(
    title = "Prevalence Rate vs Mortality Rate by Disease",
    x = "Prevalence Rate (%)",
    y = "Mortality Rate (%)",
    color = "Disease"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "none"  # Hide if too many disease types
  )
```
```{r}
#Group by country and compute averages
grouped_df <- final_data %>%
  group_by(country) %>%
  summarise(
    healthcare_access_avg = mean(healthcare_access, na.rm = TRUE),
    mortality_rate_avg = mean(mortality_rate, na.rm = TRUE)
  ) %>%
  arrange(country) %>%
  mutate(Country_Number = row_number())  # Capitalize to match exactly

#Create a label column for the legend
grouped_df <- grouped_df %>%
  mutate(Label = paste0(Country_Number, " - ", country))

#Plot
ggplot(grouped_df, aes(x = healthcare_access_avg, y = mortality_rate_avg)) +
  geom_point(aes(color = factor(Country_Number)), size = 4) +
  geom_text(aes(label = Country_Number), vjust = -1, size = 3.5) +
  scale_color_manual(
    values = scales::hue_pal()(nrow(grouped_df)),
    labels = grouped_df$Label
  ) +
  labs(
    title = "Healthcare Access vs Mortality Rate by Country",
    x = "Healthcare Access (%)",
    y = "Mortality Rate (%)",
    color = "Country (No - Name)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10)
  )
```
Observations:
No Strong Direct Relationship Between healthcare_access and mortality_rate
Among the 8 countries analyzed, higher healthcare_access does not consistently correspond to lower mortality_rate. This suggests that access alone might not be the sole factor driving mortality — quality, distribution, or utilization of healthcare services could be influential.

USA (Country 8): Highest healthcare_access, But Elevated mortality_rate
Despite having the highest average healthcare access, the USA still shows a relatively high mortality_rate. This could imply inefficiencies in the healthcare system, high costs (average_treatment_cost_usd), or disparities in access across subgroups.

China (Country 2): Lowest mortality_rate Despite Mid-Level Access
China displays the lowest average mortality_rate, despite only moderate healthcare_access. This might reflect effective public health measures, better availability_of_vaccines_treatment, or a population profile with fewer comorbidities.

Argentina (Country 1): Low healthcare_access, High mortality_rate
Argentina ranks lower in healthcare accessibility and higher in mortality, which could signal gaps in health infrastructure, limited hospital_beds_per_1000, or lower availability of essential treatments.

Russia (Country 5) and South Africa (Country 6): High Access, Yet High Mortality
These countries show that even with comparatively high healthcare_access, their mortality_rate remains elevated. This could be due to challenges like high incidence_rate of chronic or infectious diseases, lower recovery_rate, or socioeconomic disparities (per_capita_income_usd, education_index).

India (Country 3): Moderate Access, Moderate Mortality
India maintains a balanced profile, where moderate healthcare_access correlates with an acceptable mortality_rate. This might reflect efficient resource use, lower average_treatment_cost_usd, or stronger community-level care.
```{r}
library(gridExtra)
library(grid) 
library(ggplot2)
library(dplyr)
library(scales)
library(RColorBrewer)  # For color palettes


# Filter to the 8 countries of interest and years 2020–2024
target_countries <- c("Argentina", "China", "India", "Nigeria", "Russia", "South Africa", "UK", "USA")
df_filtered <- final_data %>%
  filter(country %in% target_countries, year >= 2020 & year <= 2024)

# Find top disease per year for each country based on population_affected
top_diseases <- df_filtered %>%
  group_by(country, year, disease_name) %>%
  summarise(total_affected = sum(population_affected, na.rm = TRUE), .groups = 'drop') %>%
  group_by(country, year) %>%
  slice_max(order_by = total_affected, n = 1, with_ties = FALSE) %>%
  ungroup()

# Create individual plots
plots <- lapply(target_countries, function(cntry) {
  plot_data <- top_diseases %>% filter(country == cntry)
  
  ggplot(plot_data, aes(x = factor(year), y = total_affected, fill = disease_name)) +
    geom_bar(stat = "identity") +
    
    # Data labels in the center, vertical
    geom_text(aes(label = disease_name),
              angle = 90, vjust = 0.5, hjust = 0.5, size = 3, color = "black") +

    # Apply a color palette
    scale_fill_brewer(palette = "Set2") +

    # Format Y-axis as K
    scale_y_continuous(labels = function(x) paste0(x / 1000, "K")) +

    labs(title = cntry, x = "Year", y = "Population Affected") +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      axis.title.x = element_text(size = 8),
      axis.title.y = element_text(size = 8),
      legend.position = "none",
      legend.key.size = unit(0.4, "cm")
    )
})

# Draw the grid layout
grid.newpage()
grid.draw(arrangeGrob(grobs = plots, ncol = 4,
                      top = textGrob("Top Disease per Year by Population Affected (2020–2024)",
                                     gp = gpar(fontsize = 14, fontface = "bold"))))

```
 Country-Wise Observations
🇦🇷 Argentina
Each year shows a different disease as most widespread: from Ebola to Measles.

This variety might suggest changing public health priorities or seasonal outbreaks.

🇨🇳 China
Cancer dominated two years (2022–2023), and Asthma rose sharply in 2024.

The upward trend in Asthma cases may indicate environmental or air quality concerns.

🇮🇳 India
A strong presence of neurodegenerative and chronic diseases like Parkinson’s, Alzheimer’s, and HIV/AIDS.

This could reflect an aging population and increasing non-communicable disease burden.

🇳🇬 Nigeria
Leprosy dominates 3 out of 5 years — a rare but regionally persistent disease.

Tuberculosis and Polio also feature, reflecting infectious disease challenges in the region.

🇷🇺 Russia
Consistently high cases across Zika, Hepatitis, Polio, and Tuberculosis.

The 2024 rise in Tuberculosis may signal an outbreak or declining control efforts.

🇿🇦 South Africa
Highly variable but HIV/AIDS and Diabetes dominate recent years.

Shift from infectious to chronic conditions in 2023–2024 may reflect transitioning health dynamics.

🇬🇧 UK
HIV/AIDS leads in multiple years, with COVID-19 peaking in 2021.

By 2024, Influenza emerges — possible return of seasonal flu after pandemic suppression.

🇺🇸 USA
Zika and Hepatitis peak in 2021–2022, followed by consistent Measles cases.

Recurring Measles in 2023–2024 could hint at vaccination gaps or hesitancy.

Cross-Country Insights
HIV/AIDS is a leading concern in India, UK, South Africa — suggesting persistent burden.

Infectious diseases like Polio, Leprosy, and Tuberculosis dominate in Nigeria and Russia.

Chronic and neurological diseases (e.g., Parkinson's, Alzheimer’s) are rising in India and Argentina.

Countries like the USA and UK see resurgence in vaccine-preventable diseases, possibly post-COVID.